---
title: "World Series Home Field"
author: "Hongyu Dai"
date: "2021/10/2"
output: 
  html_document:
    code_folding: hide
    toc: yes
    number_sections: yes
    toc_depth: 3
    toc_float: yes
---

# Introduction

**If home field advantage exists, how much of an impact does it have on winning the world series?**

The home field advantage is the edge which a team may have when playing a game at its home stadium. For example, it is the edge the Braves may have over the Yankees when the head-to-head match-up is in Atlanta. It is the advantage the Yankees may have when the head-to-head match-up is in New York.

The World Series is a first-to-4-wins match-up between the champions of the American and National Leagues of Major League Baseball. 

In this blog, I am going to use simulation and analytic methods to compare the probability of winning the World Series with and without home field advantage.

# Assumption

1. The Braves and the Yankees are teams competing in the World Series.

2. The results of each game are independent of each other

3. The winning team must win the last game

4. The game rules are 7 games 4 wins

# Setup

1. Braves's home field is Atlanta, and Yankees's home field is New York.

2. The sequence of game locations is `NYC, NYC, ATL, ATL, ATL, NYC, NYC`.

3. The table below has the two possible schedules for each game of the series. (NYC = New York City, ATL = Atlanta)

| Overall advantage | Game 1 | Game 2 | Game 3 | Game 4 | Game 5 | Game 6 | Game 7 |
|:-----------------:|:------:|:------:|:------:|:------:|:------:|:------:|:------:|
|      Braves       |  ATL   |  ATL   |  NYC   |  NYC   |  NYC   |  ATL   |  ATL   |
|      Yankees      |  NYC   |  NYC   |  ATL   |  ATL   |  ATL   |  NYC   |  NYC   |

4. Let *P*<sub>*B*</sub> be the probability that the Braves win a single head-to-head match-up with the Yankees, under the assumption that home field advantage doesn’t exist. Let *P*<sub>*B*</sub><sup>*H*</sup> denote the probability that the Braves win a single head-to-head match-up with the Yankees as the home team (H for home). Let *P*<sub>*B*</sub><sup>*A*</sup> denote the probability that the Braves win a single head-to-head match-up with the away team (A for away).

| Game location |   No advantage    | Advantage                                                            |
|:-------------:|:-----------------:|:---------------------------------------------------------------------|
|      ATL      | *P*<sub>*B*</sub> | *P*<sub>*B*</sub><sup>*H*</sup> = *P*<sub>*B*</sub> \* 1.1           |
|      NYC      | *P*<sub>*B*</sub> | *P*<sub>*B*</sub><sup>*A*</sup> = 1 − (1 − *P*<sub>*B*</sub>) \* 1.1 |

**Note**:

(The advantage factor in *P*<sub>*B*</sub><sup>*H*</sup> and *P*<sub>*B*</sub><sup>*A*</sup> is the 1.1 multiplier that results in a 10% increase for the home team. Generate a plot to answer this question.)

3. Load Packages

```{r}
library(dplyr)
library(data.table)
library(ggplot2)
```


# Questions & Answer:

## Compute analytically the probability that the Braves win the world series. Calculate the probability with and without home field advantage when *P*<sub>*B*</sub> = 0.55. What is the difference in probabilities?


```{r}
# Set probability
pb <- 0.55
advantage_multiplier <- 1.1  #Set = 1 for no advantage
pbh <- pb*advantage_multiplier
pba <- 1 - (1 - pb)*advantage_multiplier


# 1) With home field advantage
## Get all possible outcomes
apo <- fread("all-possible-world-series-outcomes.csv")

## Home field indicator
hfi <- c(0,0,1,1,1,0,0)  #{NYC, NYC, ATL, ATL, ATL, NYC, NYC}

## Calculate the probability of each possible outcome
apo[, p := NA_real_] # Initialize new column in apo to store prob
for(i in 1:nrow(apo)){
  prob_game <- rep(1, 7)
  for(j in 1:7){
    p_win <- ifelse(hfi[j], pbh, pba)
    prob_game[j] <- case_when(
        apo[i,j,with=FALSE] == "W" ~ p_win
      , apo[i,j,with=FALSE] == "L" ~ 1 - p_win
      , TRUE ~ 1
    )
  }
  apo[i, p := prod(prob_game)] # Data.table syntax
}

## Sanity check: does sum(p) == 1?
#apo[, sum(p)] # This is data.table notation

## Probability of overall World Series outcomes
apo[, sum(p), overall_outcome]

## Braves’ probability of winning
ans1 <- as.numeric(apo[, sum(p), overall_outcome][1,2])
ans1

# 2) Without home field advantage
ANS1 <- pnbinom(3, 4, pb) 
```

+ For `NYC, NYC, ATL, ATL, ATL, NYC, NYC` sequence, With and without home advantage, the Braves' probability of winning is 0.604221 and 0.6082878 respectively. 

+ The probability of winning without home advantage is slightly greater than that of winning with home advantage. It can be seen that home field advantage do have an impact on winning the world series.



## Calculate the same probabilities as the previous question by simulation.

```{r}
set.seed(10)
times=10000  # set simulation times

# 1) With home field advantage
result2 <- vector()
for (i in 1:times){
  game2_result <- vector()
  for (j in 1:7) { # j: number of game in one simulation
    p_win2 <- ifelse(hfi[j], pbh, pba) #p_win2: winning probability of each game
    game2_result[j] = rbinom(1, 1, p_win2) #Generate game results based on p_win2. 1 is considered as a win, 0 as a loss
    if (j >= 4) {
      if ((sum(game2_result, na.rm = TRUE) == 4) || (sum(1 - game2_result, na.rm = TRUE) == 4))  {
        break  #One series ends when one side has 4 wins, i.e. one simulation ends
      }
    }
  }
  #print(game2_result)  # see one series game result
  result2[i]=ifelse(sum(game2_result)==4,1,0) # Final results of all simulations
}
(ans2 <- mean(result2))
```

```{r}
set.seed(20)
# 2) Without home field advantage
RESULT2 <- vector()
for (i in 1:times){
  game2_result <- vector()
  for (j in 1:7) { 
    game2_result[j] = rbinom(1, 1, pb)
    if (j >= 4) {
      if ((sum(game2_result, na.rm = TRUE) == 4) || (sum(1 - game2_result, na.rm = TRUE) == 4))  {
        break 
      }
    }
  }
  RESULT2[i]=ifelse(sum(game2_result)==4,1,0)
}
(ANS2 <- mean(RESULT2))
```

By simulation solution, we can also get the answer. With and without home advantage, the Braves' probability of winning is 0.6072 and 0.6002 respectively. 

## What is the absolute and relative error for your simulation in the previous question?

According to abs_error=$|p̂−p|$, rel_error = $|p̂−p|/p$, we can get the answer as below: 

```{r}
# 1) With home field advantage
(abs_err_withHFA <- abs(ans2-ans1))
(rel_err_withHFA <- abs(ans2-ans1)/ans1)

#2) Without home field advantage
(abs_err_NoHFA <- abs(ANS2-ANS1))
(rel_err_NoHFA <- abs(ANS2-ANS1)/ANS1)
```

## Does the difference in probabilities (with vs without home field advantage) depend on *P*<sub>*B*</sub>? (Generate a plot to answer this question.)

```{r}
PB <- seq(0.5,1,by = 0.01)

## 1) With home field advantage
apo[, p := NA_real_] 
ans4 <- vector()
for (k in seq_along(PB)){
  advantage_multiplier <- 1.1  #Set = 1 for no advantage
  PBH <- PB[k]*advantage_multiplier
  PBA <- 1 - (1 - PB[k])*advantage_multiplier
  for(i in 1:nrow(apo)){
    prob_game <- rep(1, 7)
    for(j in 1:7){
      p_win <- ifelse(hfi[j], PBH, PBA)
      prob_game[j] <- case_when(
          apo[i,j,with=FALSE] == "W" ~ p_win
        , apo[i,j,with=FALSE] == "L" ~ 1 - p_win
        , TRUE ~ 1
      )
    }
    apo[i, p := prod(prob_game)] 
  }
  ans4[k] <- as.numeric(apo[, sum(p), overall_outcome][1,2])
}

#2) Without home field advantage
ANS4 <- pnbinom(3, 4, PB) 
```


```{r}
# Creat a point plot
ggplot()+
  geom_point(aes(x=PB,y=ans4-ANS4))+
  labs(title = 'the difference in probabilities (with vs without home field advantage)',
       subtitle = 'when changing pb from 0.5 to 1',
       x='pb')+
  theme_bw()+
  theme(panel.border = element_blank()) 

# Find the max Coordinates
max(ans4-ANS4)
PB[which(near(ans4-ANS4,0.01775953))]
```

From the plot, we can seen that *P*<sub>*B*</sub> do have impact on the difference in probabilities (with vs without home field advantage). When `pb` was between 0.5 and 0.71, `pb` was positively correlated with `ans4-ANS4`. When `pb` increased from 0.71 to 1, `pb` was negatively correlated with `ans4-ANS4`, gradually converging to 0.

## Does the difference in probabilities (with vs without home field advantage) depend on the advantage factor? 

```{r}
## 1) With home field advantage
advantage_multiplier <- seq(1,2,by = 0.01)
apo[, p := NA_real_] 
ans5 <- vector()
for (m in seq_along(advantage_multiplier)){
  PBH <- pb*advantage_multiplier[m]
  PBA <- 1 - (1 - pb)*advantage_multiplier[m]
  for(i in 1:nrow(apo)){
    prob_game <- rep(1, 7)
    for(j in 1:7){
      p_win <- ifelse(hfi[j], PBH, PBA)
      prob_game[j] <- case_when(
          apo[i,j,with=FALSE] == "W" ~ p_win
        , apo[i,j,with=FALSE] == "L" ~ 1 - p_win
        , TRUE ~ 1
      )
    }
    apo[i, p := prod(prob_game)] 
  }
  ans5[m] <- as.numeric(apo[, sum(p), overall_outcome][1,2])
}
```

```{r}
# Creat a point plot
ggplot()+
  geom_point(aes(x=advantage_multiplier,y=ans5-ANS1))+
  labs(title = 'the difference in probabilities (with vs without home field advantage)',
       subtitle = 'when changing advantage multiplier from 1 to 2',
       x='advantage multiplier')  +
  theme_bw()+
  theme(panel.border = element_blank()) 
```

From the plot above, it can be seen that, when advantage multiplier incresing from 1 to 2, the difference in probabilities (with vs without home field advantage) is descresing. So we can consider that the difference in probabilities (with vs without home field advantage) depend on the advantage factor.
