---
title: "Martingale Strategy of Roulette"
author: 'Hongyu Dai'
date: "Sept 5th, 2021"
output:
    html_document:
    code_folding: hide
    toc: yes
    number_sections: yes
    toc_depth: 3
    toc_float: yes

---

```{r global options, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, cache=TRUE)
```

# Introduction 
This blog is going to explain how to simulate the operating characteristics of the martingale betting strategy for roulette.  

The purpose of this blog is to enable students without any data science background to understand the application of the strategy.

This blog will help you understand:

+ What’s Roulette

+ What's Martingale Strategy

+ How to use the computer to simulate Martingale Strategy in Roulette

Please contact freely if you have any question. Email: hongyu.dai@vanderbilt.edu

# Background

## What’s Roulette
Roulette is a casino game, it's a game of chance in which a pocket is randomly selected. There is a wheel with 38 (or 37) pockets is spun in one direction, the pockets are colored red, black, or green. The pockets are also numbered. while marble is spun in the opposite direction as the momentum of the wheel and marble slows the marble will eventually come to rest in one of the pockets.

Gamblers may wager on several aspects of the outcome, they can wager that the ball will land in the red or odd-numbered pocket or will be a specific number. Then, the winnings are then paid to anyone who has placed a successful bet. 

For this blog, we will take 38 (double-zero American style roulette) as an example, consider wagers based on the color of the pocket. To be specific, there are 38 pockets of which 2 are green, 18 are red, and 18 are black. The payout for a bet on black (or red) is $1 for each $1 wagered. This means that if a gambler bets one dollar on red and the randomly selected pocket is red, then the gambler will get the original one-dollar wager. And an additional one dollar as winnings a losing one dollar wager. Otherwise, results in the loss of one dollar.

## What's Martingale Strategy

For Roulette, there are serval strategies. But this blog is going to introduce a classic roulette strategy called the “Martingale” strategy. And Introducing it with a a single sequence of spins {Black, Black, Red} example.

This strategy is very simple: in a betting board with an over or under bet, keep betting only on a single side (such as
press big or press small), and for each loss, multiply the number of losses by twice and continue until the gambler's press wins once, then he can get back the entire amount lost and win more than the amount he pressed the first time.

This seems like a surefire strategy. So, in order to investigate whether the Martingale strategy is really so effective, the following will use a computer to perform a simulation of the Roulette game, and analyze this strategy using computer simulations to explore its average earnings and limitations.

## Coding language and Package
In this blog, we are going to use R markdown software, and the package we needed is as follow:
```{r}
library("tidyverse")
```
 
# Simulation Method

We are going to using **3PR** Framework for simulations. **3PR** stands for **Parameters, Process, Property and Replicate**.

## Parameters

For this case, we need 4 parameters:


| Parameter | Description                     | Starting value             |
|:---------:|:--------------------------------:|:--------------------------------------:|
|   **B**   | Starting budget                 | $200                  |
|   **W**   | Winnings threshold for stopping | $300 (Starting budget + $100 winnings) |
|   **L**   | Tmax game threshold for stopping| 1000 plays               |
|   **M**   | Casino’s maximum wager          | $100                  |

**Note**: 

1) The player cannot wager more money than he/she has.

2) Some casinos have a maximum wager. If the strategy directs the player to wager more than M dollars, then the player will only wager M dollars.

3) To facilitate subsequent simulations, four initial values were assigned to them. And suppose the gambler will always press red.*

## Process
In order to simulation a series of plays, we need to simulation a single play of the martingale strategy first. And for a single play of the martingale strategy, it can be divide into two part: *Single Spin of the Wheel* and *Single Wager*. 

### A Single Spin of the Wheel

The way to recreate the spin of a roulette wheel is to use the sample command the first input is a vector of objects the second input is the number of objects one wants to randomly select in the roulette setting:

```{r}
single_spin <- function(){
  possible_outcomes <- c(rep("red",18), rep("black",18), rep("green",2))
  sample(possible_outcomes, 1)
}
```

### A Single Wager

In order to create a single wager function, we need to know four information: current budget, previous outcome, previous wager, max wager. With these four pieces of information as input, the function will return the martingale wager:

```{r}
martingale_wager <- function(
  current_budget, 
  previous_outcome, 
  previous_wager, 
  max_wager
){
  if(previous_outcome == "red") return(1)
  min(2*previous_wager, max_wager, current_budget)
}
```

**Explanation**: 

+ if previous outcome equal to red, then the wager will return to 1, which is one dollar. If the previous outcome is not red, then the wager will return to the second parameter, which is the minimum of the maximum wager the gambler's current budget or twice the previous loss. This is Martingale strategy we want.

### One Play

By combining the above two functions, we can simulate a single roulette game using a martingale strategy. For this function: we just need to input its the previous ledger entry and the maximum wager parameter, then we can get the out result.

```{r}
one_play <- function(previous_ledger_entry, max_wager){
  #first, we creates a copy of the previous ledger entry. The new object is called 'out' and it will be the output object from the function because the output object is a ledger entry itself it makes sense to use the input object as a template.
  out <- previous_ledger_entry
  
  #second, immediately updates the values for game index and starting budget. 
  out[1, "game_index"] <- previous_ledger_entry[1, "game_index"] + 1
  out[1, "starting_budget"] <- previous_ledger_entry[1, "ending_budget"]
  
  #third, the martingale wager function is used to calculate the wager indicated by the strategy
  out[1, "wager"] <- martingale_wager(
    previous_wager = previous_ledger_entry[1, "wager"]
    , previous_outcome = previous_ledger_entry[1, "outcome"]
    , max_wager = max_wager
    , current_budget = out[1, "starting_budget"]
  )
  
  #then, the outcome of the roulette game is generated from the single spin function
  out[1, "outcome"] <- single_spin()
  
  #the ending budget is calculated by either adding or subtracting the wager from the starting budget. 
  out[1, "ending_budget"] <- out[1, "starting_budget"] + 
    ifelse(out[1, "outcome"] == "red", +1, -1)*out[1, "wager"]
  
  #with the new ledger entry complete the final code chunk returns out.
  return(out)
}
```


### Stopping Rules

In addition, we also need to set up Stopping Rules to decide when to stop roulette. After all, the game can't last forever. 

We assume a gambler will use martingale strategy and play until：

+ the gambler has W dollars

+ the gambler goes bankrupt

+ the gambler completes L wagers (or plays)

So, the above stopping rules can be summarized as:

+ ending budget <= 0

+ game index >= max plays

+ ending budget >= winning threshould

```{r}
stopping_rule <- function(
  ledger_entry
  , winning_threshold
){
  ending_budget <- ledger_entry[1, "ending_budget"]
  if(ending_budget <= 0) return(TRUE)
  if(ending_budget >= winning_threshold) return(TRUE)
  FALSE
}
```

**Note:**

1. In order to facilitate the subsequent coding of the **one_series_play** function, here we combine the three parameters of ending budget, game index, and winning threshould into a **ledger_entry**. The details of **ledger_entry** will be described below.

2. After entering the corresponding data into the evaluation by stopping rule, if the output is True, it means the gambler should stop, otherwise the gambler should continue to play.


### One Series Play

Our goal is to create a long series of plays by using this method. In order to achieve this, we are going to create a ledger, where each row corresponds to a single game. 

The columns in the ledger include the budget at the start of the individual play or game, it also includes a column for the wager placed for that game, a column for the outcome of the roulette spin, and a column to record the budget after the wins and losses are assessed. The ledger records the information from the subsequent plays and the wagers change according to the strategy and the ending budgets change according to the luck of the player.

Among them, the result of each game is calculated by **single_spin()** function. The ending budget would be updated appropriately to reflect the wins and losses of the gamble. The ledger plus the maximum wager parameter provide all the needed information in order to mimic a series of future roulette plays.

Because **one_ play()** function cannot set the starting value. Therefore, we choose 0 as the initial value of the game index and set **wager = 1, outcome = red**. Under such conditions, the **ending_budget** of **game_index=0** is the **starting_budget** of **game_index=1**, which can be applied to all subsequent ledgers.

Thus, we can design the code needed to complete One Series Play function:

```{r}
one_series <- function(
  max_games, starting_budget, winning_threshold, max_wager
){
  # Initialize ledger
  ledger <- data.frame(
      game_index = 0:max_games
    , starting_budget = NA_integer_
    , wager = NA_integer_
    , outcome = NA_character_
    , ending_budget = NA_integer_
  )
  ledger[1, "wager"] <- 1
  ledger[1, "outcome"] <- "red"
  ledger[1, "ending_budget"] <- starting_budget
  for(i in 2:nrow(ledger)){
    #browser()
    ledger[i,] <- one_play(ledger[i-1,], max_wager)
    if(stopping_rule(ledger[i,], winning_threshold)) break
  }
  # Return non-empty portion of ledger
  ledger[2:i, ]
}
```

## Property 

One of the remaining tasks is to calculate walk out money (profit or loss) from a single series of games. The walk out money is the difference between the ending budget on the last game and the starting budget on the first:

```{r}
walk_out_money <- function(ledger){
  n <- nrow(ledger)   #n is the number of rows in the ledger
  walk_out_money <- ledger[n, "ending_budget"] - ledger[1, "starting_budget"]
  return(walk_out_money)
}
```


# Discussion

## How to calculate the average earnings of a gambler that uses the martingale strategy

By assigning the initial value to **one_ series()** function, we can get the ledger of one series play:

```{r}
result= one_series(starting_budget=200, winning_threshold =300,max_games=1000,max_wager =100 )
head(result)
```
In order to explore the effect of this strategy, we will draw line graph to learn the distribution and changes of players' earnings in a series of games, and calculate the probability of making money leaving and the average earnings:

```{r}
earnings=result[,5]-200

plot(earnings,type ="l", main = "Gambler earnings VS. Wager Number", xlab = "Wager Number", ylab = "Earnings")

mean(earnings>0) #probability of making money leaving
mean(earnings)   #average earnings
```

Through these graphs, we can observe how the gambler's earnings change with the game index, and also know the probability of making money leaving and the average earnings. This is helpful in the selection of gambling strategies.

## How does changing one parameter of the simulation affect the average revenue

Many people believe that in a casino, as long as one has enough money, one can be invincible. To investigate this view, next I will assign different values to the Starting budget parameter and observe the affect of a change in the Starting budget parameter on average revenue by drawing a graph：

```{r}
changes_start_budget <- seq(50,400,50)
changes_earning <- vector()

for (i in 1:length(changes_start_budget)){
  changes_earning [i] <- one_series( starting_budget=changes_start_budget[i], winning_threshold =300,max_games=1000,max_wager =100 )  %>% walk_out_money
}
plot(changes_start_budget,changes_earning, type = "b", xlab= "Change in starting budget", ylab = "Final Earnings")
```

By observing the above figure, we can find that there is no correlation between start budget and earnings.


## How to estimate the average number of plays before stopping.

```{r}
playnumber_before_stopping <- rep(NA, 50)
for(j in seq_along(playnumber_before_stopping)){
  playnumber_before_stopping[j] <- one_series(starting_budget=200, winning_threshold =300,max_games=1000,max_wager =100)  %>% nrow()
   
}
mean (playnumber_before_stopping)
```

# Limitation

+ Due to the randomness of computer simulation, the results of operation are unstable

+ Computer simulations are based on the premise that casinos are completely fair, and that black box events that may exist in real casinos cannot be predicted through computer simulations.

