---
title: "Log Transform"
author: "Hongyu Dai"
date: "2021/10/8"
output: 
  html_document:
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float: yes
---

# Introduction

It is common in the analysis of biological data to log transform data representing concentrations or data representing dose response.

In this blog, I am to answer a series of questions related to the transformation of data. I will use both analytic methods and simulation to answer the questions.

# Setup

1. Geometic Means = $\sqrt[n]{x_1 \cdot x_2 \cdots x_n}$. The sample size used subsequently in this blog is too large for the formula to be used in R, so I will use the logarithm first and then exponentiate to calculate the geometric mean.

2. The logarithmized variable must be greater than 0all variable in log should larger than 0.

# Part 1

## Question

1. For each distribution below, generate a figure of the PDF and CDF, and mark the mean and median in the figure.

2. For each distribution below, generate a figure of the PDF and CDF of the transformation Y = log(X) random variable. And mark the mean and median in the figure. By using simulation methods in order find the PDF and CDF of the transformation.

3. For each of the distributions below, generate 1000 samples of size 100. And calculate the geometric and arithmetic mean for each sample. Generate a scatter plot of the geometic and arithmetic sample means. Add the line of identify as a reference line.

4. Generate a histogram of the difference between the arithmetic mean and the geometric mean.

## Answer

### Distribution 1

*X* ∼ GAMMA(shape = 3, scale = 1)

1. For Gamma distribution,

$$Mean = Shape*Scale$$

```{r}
x = seq(0, 10, by = 0.01)
mean1 <- 3*1
median1 <- qgamma(0.5,shape = 3, scale = 1)
```

```{r}
pdf1 <- dgamma(x, shape = 3, scale = 1)

# plot pdf figure
plot(x,pdf1,main='Gamma Distribution\'s PDF figure',type='l')
abline(v=mean1,col='red') 
abline(v=median1,col='blue',lty=2)
legend("topright",c("mean","median"),col=c("red","blue"),lty=c(1,2),bty = "n")
```

```{r}
cdf1 <- pgamma(x, shape = 3, scale = 1)

# plot cdf figure
plot(x,cdf1,main='Gamma Distribution\'s CDF figure',type='l')
abline(v=mean1,col='red') 
abline(v=median1,col='blue',lty=2)
legend("topright",c("mean","median"),col=c("red","blue"),lty=c(1,2),bty = "n")
```

2.

```{r}
gamma2<-function(times,shape, scale,title1,title2){
X <- rgamma(times,shape, scale)
Y <- log(X)
plot(density(Y),type='l',main=title1,sub='when Y=log(X)')
abline(v=mean(Y),col='red') 
abline(v=median(Y),col='blue',lty=2)
legend("topleft",c("mean","median"),col=c("red","blue"),lty=c(1,2),bty = "n")

plot(ecdf(Y),main=title2)
abline(v=mean(Y),col='red') 
abline(v=median(Y),col='blue',lty=2)
legend("topleft",c("mean","median"),col=c("red","blue"),lty=c(1,2),bty = "n")
}

gamma2(1000,3,1,title1='GAMMA Distribution\'s PDF figure',title2='GAMMA Distribution\'s CDF figure')
```

3 & 4.

```{r}
gamma34<-function(n,size,shape, scale){

x=as.data.frame(replicate(n,rgamma(size,shape, scale)))

# arithmetic mean & ge mean
ar_mean <- colMeans(x)
ge_mean <- exp(colMeans(log(x))) 

plot(x=ar_mean,y=ge_mean,main='Arithmetic Mean VS Geometric Mean ',xlab = "Arithmetic Mean",  ylab = "Geometric Mean ")
abline(0,1,col='red')

hist(ar_mean-ge_mean,main= 'The Difference between arithmetic mean & geometric mean',xlab='Difference: Arithmetic Mean - Geometric Mean')
}

gamma34(1000,size=100,shape = 3, scale = 1)
```


### Distribution 2

*X* ∼ LOG NORMAL(*μ* =  − 1, *σ* = 1)

1. For log normal distribution,

$$Mean=exp(μ+σ^2/2)$$

$$Median=exp(μ)$$

```{r}
x2<-seq(0,5,0.01)
mean2<-exp(-1+1^2/2)
median2<-exp(-1)
```

```{r}
pdf2 <- dlnorm(x2,-1,1)

plot(x2,pdf2,main='LOG NORMAL Distribution\'s PDF figure',type='l')
abline(v=mean2,col='red') 
abline(v=median2,col='blue',lty=2)
legend("topright",c("mean","median"),col=c("red","blue"),lty=c(1,2),bty = "n")
```

```{r}
cdf2 <- plnorm(x2,-1,1)

plot(x2,cdf2,main='LOG NORMAL Distribution\'s CDF figure',type='l')
abline(v=mean2,col='red') 
abline(v=median2,col='blue',lty=2)
legend("bottomright",c("mean","median"),col=c("red","blue"),lty=c(1,2),bty = "n")
```

2.

```{r}
lnorm2<-function(times,μ,σ,title1,title2){
X <- rlnorm(times,μ,σ)
Y <- log(X)
plot(density(Y),type='l',main=title1,sub='when Y=log(X)')
abline(v=mean(Y),col='red') 
abline(v=median(Y),col='blue',lty=2)
legend("topleft",c("mean","median"),col=c("red","blue"),lty=c(1,2),bty = "n")

plot(ecdf(Y),main=title2)
abline(v=mean(Y),col='red') 
abline(v=median(Y),col='blue',lty=2)
legend("topleft",c("mean","median"),col=c("red","blue"),lty=c(1,2),bty = "n")
}

lnorm2(1000,-1,1,title1='Log NORMA Distribution\'s PDF figure',title2='Log NORMA Distribution\'s CDF figure')
```

3 & 4.

```{r}
lnorm34<-function(n,size,μ,σ){

x=as.data.frame(replicate(n,rlnorm(size,μ,σ)))

# arithmetic mean & ge mean
ar_mean <- colMeans(x)
ge_mean <- exp(colMeans(log(x))) 

plot(x=ar_mean,y=ge_mean,main='Arithmetic Mean VS Geometric Mean ',xlab = "Arithmetic Mean",  ylab = "Geometric Mean ")
abline(0,1,col='red')

hist(ar_mean-ge_mean,main= 'The Difference between arithmetic mean & geometric mean',xlab='Difference: Arithmetic Mean - Geometric Mean')
}

lnorm34(1000,size=100,μ=-1,σ=1)
```

### Distribution 3

*X* ∼ UNIFORM(0, 12)

1. For uniform distribution, 

$$Mean=\frac{a+b}{2}$$

```{r}
x3 <- seq(-2,14,0.01)
mean3 <- (0+12)/2
median3 <-qunif(0.5,0,12)
```

```{r}
pdf3 <- dunif(x3,0,12)

plot(x3,pdf3,main='UNIFORM Distribution\'s PDF figure',type='l')
abline(v=mean3,col='red') 
abline(v=median3,col='blue',lty=2)
legend("topright",c("mean","median"),col=c("red","blue"),lty=c(1,2),bty = "n")
```

```{r}
cdf3 <- punif(x3,0,12)

plot(x3,cdf3,main='UNIFORM Distribution\'s CDF figure',type='l')
abline(v=mean3,col='red') 
abline(v=median3,col='blue',lty=2)
legend("bottomright",c("mean","median"),col=c("red","blue"),lty=c(1,2),bty = "n")
```

2.

```{r}
unif2<-function(times,a,b,title1,title2){
X <- runif(times,a,b)
Y <- log(X)
plot(density(Y),type='l',main=title1,sub='when Y=log(X)')
abline(v=mean(Y),col='red') 
abline(v=median(Y),col='blue',lty=2)
legend("topleft",c("mean","median"),col=c("red","blue"),lty=c(1,2),bty = "n")

plot(ecdf(Y),main=title2)
abline(v=mean(Y),col='red') 
abline(v=median(Y),col='blue',lty=2)
legend("topleft",c("mean","median"),col=c("red","blue"),lty=c(1,2),bty = "n")
}

unif2(1000,a=0,b=12,title1='UNIFORM Distribution\'s PDF figure',title2='UNIFORM Distribution\'s CDF figure')
```

3 & 4.

```{r}
unif34<-function(n,size,a,b){

x=as.data.frame(replicate(n,runif(size,a,b)))

# arithmetic mean & ge mean
ar_mean <- colMeans(x)
ge_mean <- exp(colMeans(log(x))) 

plot(x=ar_mean,y=ge_mean,main='Arithmetic Mean VS Geometric Mean ',xlab = "Arithmetic Mean",  ylab = "Geometric Mean ")
abline(0,1,col='red')

hist(ar_mean-ge_mean,main= 'The Difference between arithmetic mean & geometric mean',xlab='Difference: Arithmetic Mean - Geometric Mean')
}

unif34(1000,size=100,a=0,b=12)
```


# Part 2 (OPTIONAL)

Show that if *X*<sub>*i*</sub> \> 0 for all *i*, then the arithmetic mean is greater than or equal to the geometric mean.

  

**Answer**

```{r}
ans2<- function(number,i){
  ar_mean = NA
  ge_mean = NA
  for (n in 1:i){
    X=sample(number,i,replace=T)
    ar_mean[n]=mean(X)
    ge_mean[n]=exp(mean(log(X)))
}
  unique(ifelse(ar_mean >=ge_mean ,"Prove","Unprove"))
}
ans2(1000,1000)
```

According to the simulation above, we can know that: if *X*<sub>*i*</sub> \> 0 for all *i*, then the arithmetic mean is greater than or equal to the geometric mean.

# Part 3

What is the correct relationship between `E[log(X)]` and `log(E[X])`? Is one always larger? Equal? Explain your answer.

**1. Analytic Answer**

Recall that:

$$e^x\geq 1+x$$

so, $$E\left[e^{Y}\right]=e^{ E(Y)}E\left[e^{Y- E(Y)}\right]\geq  e^{E(Y)} E\left[1+{Y- E(Y)}\right] = e^{E(Y)}$$.

which means:

$$e^{E(Y)}\leq E\left[e^{Y}\right]$$

Now letting $Y=\log X$, we have: 

$$e^{E(\log X)}\leq E\left[e^{\log X}\right]=E(X)$$

Now take logs of both sides, we get: 

$$E[\log (X)]\leq\log[E(X)]$$

**Conclusion** 

the relationship between `E[log(X)]` and
`log(E[X])` is that: `E[log(X)]` always less or equal than `log(E[X])`

    

**2. Simulation Answer**

```{r}
ans3<- function(number,i){
  mean_log = NA
  log_mean = NA
   diff=NA
  for (n in 1:i){
    X=sample(number,i,replace=T)
    mean_log[n]=mean(log(X))
    log_mean[n]=log(mean(X))
    diff[n]=mean_log[n]-log_mean[n]
}
  if (max(diff)<0){
    print('E[log(X)] is always smaller than log(E[X])')
  }
   if (min(diff)>0){
     print('E[log(X)] is always larger than log(E[X])')
   }
}
ans3(1000,1000)
```
 
From the simulation above, we can find out that the max `E[log(X)] - log(E[X])` is always less than 0. So `log(E[X])` is always larger.